Cosmic Microwave Background Power Spectra from Few Bit Timestreams

### 2018-08-14 digitisation.tex
Editing
Done another editing pass and adjusted the placement of the figures.

\usepackage[table,usenames,dvipsnames]{xcolor}
%\usepackage{amsmath}
%\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
%\usepackage{natbib}
\bibliographystyle{fapj}
%\usepackage{graphicx}
\usepackage{multirow}
\usepackage{soul}

@@

\begin{abstract}
Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We investigate using extreme digitisation to address these issues. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present optimal 1, 2 and 3 bit digitisation schemes and determine the degradation in temperature and polarisation power spectra caused by applying this process to the time-ordered-data (TOD). We find that 3 bit digitisation has a percent-level contribution to the map noise level. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

@@

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since 1964. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. The most prominent science goal is the discovery of the imprint of inflationary gravitational waves. Additionally studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will open up access to unprecedented insight. Moreover, through probing the relativistic number of species, the helium fraction and the neutrino mass sum, CMB experiments are a valuable counter-part to ground-based particle physics experiments.

@@

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the time-ordered-data. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their TOD in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a review of the current compression techniques employed we are sure to lose information.

In this work we present the method of extreme digitisation, which compresses a rich digital input signal and compresses it into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation observations. We find that an optimal 3-bit digitisation scheme adds as little as $\sim 2\%$ to the map noise level. While the digtisation schemes described here are primarily laid out for ground-based experiments, the results we present here should also be considered by future space-based missions, which inevitably must incorporate lossy compression.

This work is structured as follows. We detail the arising challenges in handling large TOD in \S\ref{subsec:problem}. We subsequently formulate extreme digitisation formerly in \S\ref{subsec:extremedigitisation} and lay out the framework used to test its performance in \S\ref{subsec:method}. We detail how we estimate the power spectra from simulated maps in \S\ref{subsec:psestimation}. We continue by presenting the noise induced through the dicitisation process in \S\ref{subsec:additionalnoise}. We summarize our findings in \S\ref{sec:conclusions}.

@@

The best observations sites for CMB measurements are at remote locations: in space and in the Antarctica. Experiments at these locations depend on satellite transmission. The next generation of ground-based experiments will aim to collect 2 million detector years worth of data. An experiment contributing to Stage-4 at the South Pole will face a data influx of $\sim O(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in the coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in high frequencies will go unnoticed for longer.

@@

Beyond transmission challenges, mission planning is becoming exceedingly difficult. As noted by (S4 science book) a full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to decide on Stage-4 configurations. Space-based missions must aim to carry out a similar analysis to optimise their science output. Given the shear size of TOD expected, this is not possible. We must rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages and guarantee maximal science output.

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (such as FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. While it needs to be investigated to what extent existing algorithms can be carried over, extreme digitisation has the possibility of solving planning and analysis problems.

Other science areas have demonstrated that extreme digitisation is a valuable compression technique. Jenet and Anderson (cite) explored the application of such compression to radio pulsar timing measurements with success. Recently Clearwater et al. (in prep.) have investigated the advantages of using 1 and 2 bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-wave Observatory (LIGO) performance of searches for continuous gravitational wave searches using 1 and 2 bit data.

@@

Digitisation is a lossy compression technique. However the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of input signal. A theoretical framework to obtain these levels was laid out by Max in 1978. We review the key aspects of his work relevant for us below.

Digitisation discretises an input signal by sorting it into $N$ appropriate ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output at $y_i$. A digitisation scheme is described by the number of ranges, $N$, the endpoints of these ranges, $x_k$, and the output levels, $y_k$. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$. In order to quantify the performance of a given digitisation scheme we define the distortion as

@@

which informs us that an output level $y_i$ must lie halfway between is delimiting thresholds $x_i$ and $x_{i+1}$. We gain an additional condition from equation \ref{eq:distderiv2}

@@

To progress further we have to make an assumption about the distribution of input signals, $p(x)$. For our purposes we may safely assume that ground-based CMB observations operate at low signal to noise. Furthermore we assume that the noise profile is Gaussian white noise\footnote{Please see the conclusion for a discussion of the effect of more realistic noise profiles.}, i.e. $p(x) = 1/\sqrt{2\pi} e^{-x^2/2}$. Given this assumption we can solve the problem using a numerical iterative procedure. One begins by picking $y_1$ and calculating the remaining $x_i$'s and $y_i$'s using equation \ref{eq:digitequalspacecondition}. Afterwards one observes whether this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_k$'s and $y_k$'s were chosen appropriately.

@@

to the input signal. Finally the optimal 3 bit digitisation is described by the eight-level function

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.

To create the template maps we use the healpix framework and the wealth of support available for it. We generate a realisation of I, Q and U maps based on the results of Planck 2015. The key cosmological parameters are summarised in \ref{tab:inputcosparams}.

@@

At this point we apply the digitisation schemes to the TOD. We compress the timestream into maps by averaging all hits falling into the same pixel. We produce 13 maps in total: three maps (I, Q, U) that have not undergone few-bit digitisation at the timestream level for comparison purposes, nine maps with three each corresponding to each digitisation scheme and the hitmap.

This process is carried out 7 times, producing maps with different numbers of hpp. The simulation parameters are summarised in table \ref{tab:modelparams}. The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hpp. To carry out this simulation we make use of parallelisation and the computing power provided by NERSC.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
\small
\begin{tabular}{l | c}
Parameter & Planck 2015\\
\hline
$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
$H_0$ & $67.26 \pm 0.98$\\
$n_s$ & $0.9652 \pm 0.0062$\\
$\Omega_m$ & $0.316 \pm 0.014$\\
$\sigma_8$ & $0.830 \pm 0.015 $\\
$\tau$ & $0.078 \pm 0.019$\\
$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
\end{tabular}
\tablecomments{ 
Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
} \normalsize
\end{center}
\end{table}

\begin{table}[tbh] % transpose and include hpp!
\begin{center}
\caption{\label{tab:modelparams} Assumed Survey Parameters}
\small
\begin{tabular}{l | c}
Parameter & Value\\
\hline
$\mathrm{NSIDE}$ &$4096$\\
$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
\multirow{2}{*}{$\mathrm{hpp}$} & $\left(1, 10, 100 \right) \times 800$\\
& $\left( 1, 10, 100 \right) \times 1024000$ \\
\end{tabular}
\tablecomments{ 
Parameters used in the simulated observation. The RA speed is adjusted to match the desired hpp.
} \normalsize
\end{center}
\end{table}

@@

We use PolSpice to compute the TT, EE and BB power spectra of the reconstructed maps I, Q, U maps. When doing so we apodise the observed skypatch using a cosine mask to minimise cut sky effects on the spectra. Please see \ref{tab:polspiceparams} for an overview of the parameters used in this process.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:polspiceparams} PolSpice Parameters}
\small
\begin{tabular}{c c c c}
weights & $\sigma_{\mathrm{APOD}} [\mathrm{deg^2}]$ & apodistation type & polarisation \\
\hline
cosine mask & $\sqrt{600}/2$ & cosine & Yes \\

@@

To normalise the the obtained power spectra we focus on the $\sim 10^8 \mathrm{hpp}$ maps. We normalise each power spectrum originating from few bit TOD against its corresponding 64bit TOD counterpart. We place a lower limit on the normalisation window in multipole space by considering the size of the observed patch with

@@

\[ \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \leq 10 \]

@@

we close the normalisation window the first instance the above condition is met. The normalisation constants for each digitisation scheme and channel obtained through this way are then applied to all lower hpp simulations. A sample of the obtained power spectra are shown in figures \ref{fig:ttzoom}, \ref{fig:eezoom}, and \ref{fig:bbzoom}.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
  \caption[Current ]{
  $\mathrm{TT}$ power spectrum reconstructed from a $\sim 10^8\mathrm{hpp}$ map.
           \label{fig:ttzoom}
  }
\end{figure}

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
  \caption[Current ]{
  $\mathrm{EE}$ power spectrum reconstructed from a $\sim 10^8\mathrm{hpp}$ map.
\label{fig:eezoom}
}
\end{figure}

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
  \caption[Current ]{
  $\mathrm{BB}$ power spectrum reconstructed from a $\sim 10^8\mathrm{hpp}$ map.
\label{fig:bbzoom}
}
\end{figure}

@@

To quantify how much the quality of the power spectra suffers from the digitisation process we compare the map noise levels inferred from the power spectra. These are put into the context of the map noise levels deduced from the 64bit TOD (control) power spectra. We formulate

@@

where $\sigma_{\mathrm{map}}^{\mathrm{C}}$ is the map noise level of the control power spectra and $\sigma_{\mathrm{map}}^{\mathrm{D}}$ is the map noise level obtained from the power spectra originating from extremely digitised TOD. We assume that the digitisation process results in adding a constant noise term to the power spectrum. If we are dominated by noise we can write

@@

The value of $l$ at which we can safely assume to be noise dominated and apply the above framework varies between simulated observations of different hpp and TT, EE and BB power spectra. The range considered involves any datapoints at multipole moments larger than the last point at which the detector noise is at least an order of magnitude larger than the template power spectrum, i.e.

@@

Before analysing $C_l^X/C_l^N$ we rebin the power spectra to $\Delta l = 123$. This guarantees that the points in the noise tail are independent of one another, allowing us to extract an uncertainty for the above quantity. Plots for $C_l^X/C_l^N$ are shown in figures \ref{fig:clxclntt}, \ref{fig:clxclnee}, and \ref{fig:clxclnbb} for $TT$, $EE$, and $BB$ respectively.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttratio.png}
  \caption[Current ]{
  Calculated ratio of $C_l^X/C_l^N$ of the rebinned $\mathrm{TT}$ power spectrum. The vertical black line indicates from which point onwards data is used in calculating the equivalent noise level.
           \label{fig:clxclntt}
  }
\end{figure}

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/eeratio.png}
  \caption[Current ]{
  Calculated ratio of $C_l^X/C_l^N$ of the rebinned $\mathrm{EE}$ power spectrum.
\label{fig:clxclnee}
}
\end{figure}

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/bbratio.png}
  \caption[Current ]{
  Calculated ratio of $C_l^X/C_l^N$ of the rebinned $\mathrm{BB}$ power spectrum.
\label{fig:clxclnbb}
}
\end{figure}

The deduced additional noise for 1, 2 and 3 bit digitisation schemes are shown with respect to the hits per pixel in the maps in figure \ref{fig:extranoise}. We would like to point out three key results. Firstly, 3 Bit digitisation performs the best, followed by 2 Bit and finally 1 Bit digitisation. This is what we expect, given that with each additional bit we retain more information. Secondly, for a fixed detector noise level, the additional percentage to the map noise level due to digitisation is independent of the number of hpp. The added noise therefore scales in the same fashion as the map noise level with the number of hpp, given a fixed detector noise level. Lastly, the added noise levels are astonishingly low. Keeping in mind that CMB detector sensitivity improves in steps of order of magnitude every few years adding an extra percent-level noise term does not deteriorate the results appreciably. This is impressive, given that the use of an optimal 3 bit digitisation scheme will save approximately an order of magnitude in TOD volume.

@@

Extreme digitisation is a viable lossy compression technique when dealing with low signal to noise, large datasets with a signal that is slowly varying with respect to the sampling rate. Under these conditions we have a locally flat, low signal, on top of which we add many noise realisations. In the limit of many samples these allow us to reconstruct the signal well. Furthermore a small signal will prevent saturation of the output, i.e. the inability of the digitised output to communicate any information where in the range $\infty$ to $x_N$ the input signal lies.

@@

It should be laid out how this changes when moving to a more realistic noise profile. We do not expect this to alter the practicality of our results - even if a moderate change in the noise profile doubles the additional percentage to the map noise level extreme digitisation is still practical. Ideas on how to deal with 1/f noise, e.g. chunking of the data before applying extreme digitisation have already been investigated by Planck. These thoughts should be considered when designing the compression schemes of future space-based CMB missions, which will be unable to recover their full data with latency, but rather entirely rely on the transmitted data.

@@

\bibliography{digitisation}

>>>>

\usepackage[table,usenames,dvipsnames]{xcolor}
%\usepackage{amsmath}
\usepackage{subfigure}
\usepackage[backref,breaklinks,colorlinks,citecolor=blue]{hyperref}
\usepackage{natbib}
%\usepackage{natbib}
\bibliographystyle{fapj}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{soul}

@@

\begin{abstract}
Observations of the Cosmic Microwave Background (CMB) are of immense value to modern cosmology. However, future CMB experiments must confront challenges in mission planning, hardware and analysis that arise from the sheer size of the time-ordered-data (TOD) being recorded. These challenges are particularly significant for Antarctic and satellite experiments which depend on satellite links to transmit their data. We investigate using extreme digitisation to address these issues. Unlike lossless compression, extreme digitisation introduces additional noise into the data. We present optimal 1, 2 and 3 bit digitisation schemes and determine the degradation in temperature and polarisation power spectra caused by applying this process to the TOD. We find that 3 bit digitisation has a percent-level contribution to the map noise level. This is impressive considering that it would reduce the data volume by an order of magnitude. We argue that extreme digitisation is a promising strategy for upcoming experiments.
\end{abstract}

@@

Observations of the Cosmic Microwave Background (CMB) have played a key role in physics since 1964. Current and future CMB experiments will continue to deliver new insights by studying the temperature and polarisation information contained in the CMB. These will put tight constraints on cosmological models. The most prominent science goal is the discovery of the imprint of inflationary gravitational waves. Additionally, studies of CMB lensing and the Sunyaev-Zeldovich (SZ) effects will open up access to unprecedented insight. Moreover, through probing the relativistic number of species, the helium fraction and the neutrino mass sum, CMB experiments are a valuable counter-part to ground-based particle physics experiments.

@@

A growing hurdle for experiments at remote locations are the transmission limitations of satellite links. Space-based experiments have employed a combination of lossless and lossy compression techniques, including reduced bits in the TOD. Antarctica-based experiments that transmit a portion of their data via a satellite link have downsampled their data in the past to meet their telemetry requirements. They have yet to exploit few bit digitisation of the TOD. As we approach the next generation of ground-based experiments, Stage-4, and the launch of a new generation of space-based missions (liteBIRD, PIXIE, COrE+), we must treat the transmission bottleneck carefully. Without a review of the current compression techniques employed we are sure to lose information.

In this work we present the method of extreme digitisation, which compresses a rich digital input signal and compresses it into a few bits. We apply extreme digitisation to the TOD and detail its effect on temperature and polarisation power spectra. We find that an optimal 3-bit digitisation scheme adds as little as $< 2\%$ to the map noise level. While the digtisation schemes described here are primarily laid out for ground-based experiments, the results we present here should also be considered by future space-based missions, which inevitably must incorporate lossy compression.

This work is structured as follows. We detail the arising challenges in handling large TOD in \S\ref{subsec:problem}. We subsequently formulate extreme digitisation in \S\ref{subsec:extremedigitisation} and lay out the framework used to test its performance in \S\ref{subsec:method}. The details of the power spectrum estimation used are laid out in \S\ref{subsec:psestimation}. We continue by presenting the noise induced through the dicitisation process in \S\ref{subsec:additionalnoise} and summarise our findings in \S\ref{sec:conclusions}.

@@

The best observations sites for CMB measurements are at remote locations: in space and in the Antarctica. Experiments at these locations depend on satellite transmission. The next generation of ground-based experiments will aim to collect 2 million detector years of data. An experiment contributing to Stage-4 at the South Pole will face a data influx of $\sim \mathrm{O}(10)\mathrm{Tb/d}$. However, the current transmission allocation for SPT3G is at $150\mathrm{Gb/d}$, which will likely only see a moderate increase in coming years. The transmission bottleneck is currently overcome by recovering the full data on hard drives with some latency and by transmitting a downsampled version of the data. The downsampling process loses high frequency information. Going into Stage-4 we anticipate that compression rates for transmission must increase by an order of magnitude. Continued use of downsampling will narrow the information window decisively - prohibiting high multipole moment science to be carried out on the transmitted dataset. This also means that any potential faults or errors in the experiment that only become visible in high frequencies will go unnoticed for longer.

@@

Beyond transmission challenges, mission planning is becoming exceedingly difficult. As noted by (S4 science book) a full simulation of TOD over the entire parameter space of detection scenarios for numerous set-ups is the desired way to decide on Stage-4 configurations. Space-based missions must aim to carry out a similar analysis to optimise their science output. Given the shear size of TOD expected, this is not possible. We must rely on different planning strategies or aim to reduce the size of the TOD in order to maximise the productivity of planning and development stages and guarantee scientific excellence.

@@

Extreme Digitisation would tackle the challenges mentioned above by reducing the size of the TOD by an order of magnitude. Together with already established lossless compression techniques (e.g. FLAC, run-length coding, Huffman coding, etc.) this will directly tackle transmission hurdles. While it needs to be investigated to what extent existing algorithms can be carried over, extreme digitisation has the potential of solving planning and analysis problems.

Other science areas have demonstrated that extreme digitisation is a valuable compression technique. Jenet and Anderson (cite) explored the application of such compression to radio pulsar timing measurements with success. Recently Clearwater et al. (in prep.) have investigated the advantages of using 1 and 2 bit data when searching for continuous gravitational waves using the Laser Interferometer Gravitational-wave Observatory (LIGO).

@@

Digitisation is a lossy compression technique. However the induced noise depends on the number of bits used, the digitisation thresholds, and the output levels chosen. To minimise the noise induced through this process one must know the nature of the input signal. A theoretical framework to obtain these levels was laid out by Max in 1978. We review the key aspects of his work relevant for us below.

Digitisation discretises an input signal by sorting it into $N$ appropriate ranges, such that an input between $x_i$ and $x_{i+1}$ produces an output at $y_i$. A digitisation scheme is described by the number of ranges, $N$, the endpoints of these ranges, $x_i$, and the output levels, $y_i$. Conventionally one chooses $x_{1} = -\infty$ and $x_{N+1} = \infty$. In order to quantify the performance of a given digitisation scheme we define the distortion as

@@

which informs us that an output level $y_i$ must lie halfway between its delimiting thresholds $x_i$ and $x_{i+1}$. We gain an additional condition from equation \ref{eq:distderiv2}

@@

To progress further we have to make an assumption about the distribution of input signals, $p(x)$. For our purposes we assume that CMB observations operate at low signal to noise. Furthermore we assume that the noise profile is Gaussian white noise\footnote{Please see the conclusion for a discussion of the effect of more realistic noise profiles.}, i.e. $p(x) = 1/\sqrt{2\pi} e^{-x^2/2}$. Given this assumption we can solve the problem using a numerical iterative procedure. One begins by picking $y_1$ and calculating the remaining $x_i$'s and $y_i$'s using equation \ref{eq:digitequalspacecondition}. Afterwards one observes whether this choice of values satisfy the conditions given by equation \ref{eq:digitareacondition}. If that is the case, the $x_i$'s and $y_i$'s were chosen appropriately.

@@

where $\sigma$ is the standard deviation of the input signal. Finally the optimal 3 bit digitisation is described by the eight-level function

@@

To investigate the performance of the derived digitisation schemes we simulate many scans over CMB template maps at the timestream level. Each scan is performed by a single detector. We obtain control maps that use 64bit TOD and maps that have undergone 1, 2 and 3 bit digitisation at the timestream level. We calculate the temperature and polarisation power spectra of each map and determine the additional noise induced through the extreme digitisation process.

To create the template maps we use the healpix framework and the wealth of support available for it. We generate a realisation of I, Q and U maps based on the results of Planck 2015. The key cosmological parameters are summarised in table \ref{tab:inputcosparams}.

@@

At this point we apply the digitisation schemes to the TOD. We compress the timestream into maps by averaging all hits falling into the same pixel. We produce 12 CMB maps in total: three control maps (I, Q, U) and nine maps with three each obtained from each digitisation scheme.

This process is carried out 6 times, producing maps with different numbers of hpp. The simulation parameters are summarised in table \ref{tab:modelparams}. The calculation outlined above has considerable computational requirements if we want to reach up to $\sim 10^8$ hpp. To carry out this simulation we make use of parallelisation and the computing facilities provided by NERSC.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:inputcosparams} Input Cosmological Parameters}
\small
\begin{tabular}{l | c}
Parameter & Planck 2015\\
\hline\\[-1.5ex]
$100\theta_{MC}$ & $ 1.04086 \pm 0.00048$\\
$\Omega_b h^2$ & $  0.02222 \pm 0.00023$\\
$\Omega_c h^2$ & $ 0.1199 \pm 0.0022$\\
$H_0$ & $67.26 \pm 0.98$\\
$n_s$ & $0.9652 \pm 0.0062$\\
$\Omega_m$ & $0.316 \pm 0.014$\\
$\sigma_8$ & $0.830 \pm 0.015 $\\
$\tau$ & $0.078 \pm 0.019$\\
$10^9 A_s e^{-2\tau}$ & $1.881 \pm 0.014 $\\
\end{tabular}
\tablecomments{ 
Cosmological parameters used to create the template maps. Taken from Planck 2015: Cosmological Parameters.
} \normalsize
\end{center}
\end{table}

\begin{table}[tbh] % transpose and include hpp!
\begin{center}
\caption{\label{tab:modelparams} Assumed Survey Parameters}
\small
\begin{tabular}{l | c}
Parameter & Value\\
\hline \\[-1.5ex]
$\mathrm{NSIDE}$ &$4096$\\
$f_{\mathrm{readout}} [\mathrm{Hz}]$ &$200$\\
$f_{{\mathrm{sky}}}$ & $\sim 0.014$\\
$\sigma^{\mathrm{T}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ &$500$\\
$\sigma^{\mathrm{Pol}}_{\mathrm{det}} [\mathrm{\mu K \sqrt{s}}]$ & $\sqrt{2}\times500$\\
\multirow{2}{*}{$\mathrm{hpp}$} & $\left(1, 10, 100 \right) \times 800,$\\
& $\left( 1, 10, 100 \right) \times 1024000$ \\
\end{tabular}
\tablecomments{ 
Parameters used in the simulated observation. The RA speed is adjusted to match the desired hpp.
} \normalsize
\end{center}
\end{table}

@@

We use PolSpice to compute the TT, EE and BB power spectra of the reconstructed maps I, Q, U maps. When doing so we apodise the observed skypatch using a cosine mask to minimise cut sky effects on the spectra. Please see table \ref{tab:polspiceparams} for an overview of the parameters used in this process.

\begin{table}[tbh]
\begin{center}
\caption{\label{tab:polspiceparams} PolSpice Parameters}
\small
\begin{tabular}{c c c c}
weights & $\sigma_{\mathrm{APOD}} [\mathrm{deg}]$ & apodistation type & polarisation \\
\hline
cosine mask & $\sqrt{600}/2$ & cosine & Yes \\

@@

To normalise the the obtained power spectra we focus on the $\sim 10^8 \mathrm{hpp}$ maps. We normalise each power spectrum originating from few bit TOD against its corresponding control counterpart. We place a lower limit on the normalisation window in multipole space by considering the size of the observed patch with

@@

\[ \frac{C_l^{\mathrm{S}}}{C_l^{\mathrm{N}}} \geq 10 \]

@@

we close the normalisation window the first instance the above condition is violated. The normalisation constants for each digitisation scheme and channel obtained through this way are then applied to all lower hpp simulations. A sample of the obtained power spectra are shown in figure \ref{fig:powspeczoom}.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttzoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eezoom.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbzoom.png}
  \caption[Current ]{
   Reconstructed $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. These originate from an observation with $102400000 \mathrm{hpp}$. We observe that digitisation appears to add some constant to the noise level. As one would expect $3 \mathrm{Bit}$ digitisation performs the best, followed by $2 \mathrm{Bit}$- and finally $1 \mathrm{Bit}$ digitisation.
\label{fig:powspeczoom}
}
\end{figure}

@@

To quantify how much the quality of the power spectra suffers from the digitisation process we compare the map noise levels inferred from the power spectra. These are put into the context of the map noise levels deduced from the control power spectra. We formulate

@@

where $\sigma_{\mathrm{map}}^{\mathrm{C}}$ is the map noise level of the control power spectra and $\sigma_{\mathrm{map}}^{\mathrm{D}}$ is the map noise level obtained from the power spectra originating from extremely digitised TOD. We assume that the digitisation process results in adding a constant noise term to each power spectrum. If we are dominated by noise we can write

@@

The value of $l$ at which we can safely assume to be noise dominated and apply the above framework varies between simulated observations of different hpp and temperature and polarisation power spectra. The range considered involves any datapoints at multipole moments larger than the last point at which the detector noise is at least an order of magnitude larger than the template power spectrum, i.e.

@@

Before analysing $C_l^X/C_l^N$ we rebin the power spectra to $\Delta l = 123$. This guarantees that the points in the noise tail are independent of one another, allowing us to extract an uncertainty for the above quantity. Plots for $C_l^X/C_l^N$ are shown in figure \ref{fig:clxcln}.

\begin{figure}[htb]\centering
\includegraphics[width=0.5\textwidth,clip]{Plots/ttratio.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/eeratio.png}
\includegraphics[width=0.5\textwidth,clip]{Plots/bbratio.png}
  \caption[Current ]{
  Calculated ratio of $C_l^X/C_l^N$ of the rebinned $\mathrm{TT}$, $\mathrm{EE}$, and $\mathrm{BB}$ power spectra. The vertical black line indicates from which point onwards data is used to calculate the equivalent noise level. The above power spectra originate from a $80000\mathrm{hpp}$ map. Increasing the hpp in the map shifts the plateau to the right.
\label{fig:clxcln}
}
\end{figure}

The deduced additional noise for 1, 2 and 3 bit digitisation schemes are shown with respect to the hits per pixel in the maps in figure \ref{fig:extranoise}. We would like to point out three key results. Firstly, 3 Bit digitisation performs the best, followed by 2 Bit- and finally 1 Bit digitisation. This is what we expect, given that with each additional bit we retain more information. Secondly, for a fixed detector noise level the deterioration of the map quality scales in the same fashion as the number of hpp. We see that $\Delta \sigma / \sigma$ is independent of the hpp in the maps. Lastly, the added noise levels are astonishingly low. Keeping in mind that CMB detector sensitivity improves in steps of order of magnitude every few years, adding an extra percent-level noise term does not deteriorate the results appreciably. This is impressive, given that the use of an optimal 3 bit digitisation scheme will save approximately an order of magnitude in TOD volume.

@@

Extreme digitisation is a viable lossy compression technique when dealing with low signal to noise, large datasets with a signal that is slowly varying with respect to the sampling rate. Under these conditions we have a locally flat, low signal, on top of which we add many noise realisations. In the limit of many samples these allow us to reconstruct the signal well. Furthermore a small signal will prevent saturation of the output, i.e. the inability of the digitised output to communicate any information where in the range $\infty$ to $x_N$ an input signal lies.

@@

It should be laid out how the results changes when moving to a more realistic noise profile. We do not expect this to alter the practicality of our results - even if a different noise profile doubles the additional percentage to the map noise level extreme digitisation is still practical. Ideas on how to deal with 1/f noise, e.g. chunking of the data before applying extreme digitisation have already been investigated by Planck. These thoughts should be considered when designing the compression schemes of future space-based CMB missions, which will be unable to recover their full data with latency, but rely entirely on the transmitted data.

@@

%\bibliography{digitisation}
